---
title: "Untitled"
author: "Mathis Gheno"
date: "2024-10-29"
output: html_document
editor_options: 
  chunk_output_type: console
---

## Packages

```{r}
library(tidyverse)
library(geiger)
library(purrr)
library(MASS)
library(vegan)
library(picante)
library(doParallel)
library(parallel)
```

## Usefull function

```{r}
main_theme = theme_minimal()+
  theme(line = element_blank(), 
        axis.line = element_line(colour = "black"),
        panel.border = element_blank(),
        axis.ticks =  element_line(colour = "black"),
        axis.text.x = element_text(colour = "black", size=22, face="italic", angle = 45, vjust = 1, hjust = 1),
        axis.text.y = element_text(colour = "black", size=22, face="italic"),
        legend.title = element_text(colour = "black", size=20,
                                    hjust =0.5),
        
        legend.text = element_text(colour = "black", size=18),
        axis.title= element_text(size=28),
        strip.text = element_text(colour = "black", size=15, face ="italic"))
```

## Simulation test


### Step 3 : trait-space covariance

```{r}
compute.CVtr <- function(dispersal_breadth, CVes, niche_breadth, D){
  (matrix(c(0,0,1,1,
         0,0,1,1,
         1,1,0,0,
         1,1,0,0), nrow = 4)*exp(-dispersal_breadth) +
         matrix(c(0,0,1,0,
           0,0,0,1,
           1,0,1,0,
           0,1,0,0), nrow = 4) * CVes*exp(-niche_breadth)
         )/(D+1) -> CVtr

  diag(CVtr) = c(1,1,1,1)

  colnames(CVtr) = c("trait1", "trait2", "centriod1", "centriod2")
  return(round(CVtr,2))
}
```

This is just a matrix form of the following function : $CV_{T_iR_j} = \frac{e^{-d} +CV_{E_jS_j}e^{-n_i}}{1+D_S}$. How those this work : 

- $CV_{T_iR_j}$ is the species-levels covariance. It is a covariance matrix that define the covariance between each traits ($T_i$) and each range centroid position $S_j$ (in other term : the **true mean position of each species in space**). This matrix could be estimated from real data because it define the relation between traits-traits, traits-realized niche and realized niche-realized niche. By "default" trait1-realized niche1 and trait2-realized niche2 are perfectly correlated if no process influence them. But when dispersal and fundamental niche are added to the equation, this isn't true any more.

-  The species-levels covariance matrix $CV_{T_iR_j}$ is result of the community level covariance matrix $CV_{E_jS_j}$ which is covariance of environmental variables and spatial dimensions and  by dispersal strength $e^{-d}$ and the environmental filtering (fundamental niche)  $e^{-n_i}$. Dispersal and fundamental niche only affect the traits-realized niche covariance (not traits-traits or realized niche-realized niche)
-   When there is a strong dispersal (d is big) $e^{-d} \rightarrow 0$, dispersal have no effect on traits and realized niche co-variances (dispersal isn't a limitation), . The opposite is true when there is a low dispersal
-   When fundamental niche are large (n is big) $e^{-n} \rightarrow 0$, niche have no effect on traits and realized niche co-variances (dispersal isn't a limitation). The opposite is true when fundamental niche are narrow

For exemple trait1-realized niche1 can be less corralted because of wider niche breath 
For exemple trait2-realized niche1 can be corralted because of stanger dispersal

### Step 4 : Generate regional phylogeny
```{r}
#phylo_tree <- sim.bdtree(b = 0.1, d = 0, n = num_species)  # Phylogeny of species
#plot(phylo_tree)
```

### Step 5 : Generate species traits (environmental optima)
```{r}

simulate.characher.evolution <- function(phylo_tree, CVtr){
  
  traits_and_centroid <- sim.char(phy = phylo_tree, par = CVtr, nsim = 1, model = "BM")  # Brownian motion trait evolution
  
  traits_and_centroid <- as.data.frame(traits_and_centroid)
  colnames(traits_and_centroid) <- c("trait1", "trait2", "centriod1", "centriod2")
  
  traits_and_centroid <- vegan::decostand(traits_and_centroid, method = "standardize")
  
  return(traits_and_centroid)
}
```

### Step 6 : ramdomly assing abundance from a log normal distribution
```{r}
assing.abundance <- . %>%
  mutate(abund = rpois(n(), exp(rnorm(n = n(), 5, 2))),
         realtive_abund= abund / sum(abund))
```
### Step 7 : Samples
```{r}
condition.and.position.samples<- function(num_sites, CVes){
  
  spatial_env <- MASS::mvrnorm(n = num_sites, mu = rep(0,4), Sigma = CVes)
  spatial_env <- as.data.frame(spatial_env)%>%
    vegan::decostand(method = "standardize") 
  colnames(spatial_env) <-  c("dim1", "dim2", "env1", "env2")
  
  return(spatial_env)
  }
```


### Step 8 : probability of species incidence

```{r}

create.community.matrix.probability <- function(spatial_env, traits_and_centroid){
  
 spatial_env %>%   
  rownames_to_column("sample") %>%
  rowwise() %>%
  mutate(species_probs = list(
    traits_and_centroid %>%
      rownames_to_column("species") %>%
      mutate(

        # Spatial coordinates and range centroid
        spatial_distance = sqrt((centriod1 - dim1)^2 + (centriod2 - dim2)^2),
        dispersal_prob = dnorm(spatial_distance, sqrt(dispersal_breadth)),
        
        # Environmental variables and environmental optima
        env_distance = sqrt((trait1 - env1)^2 + (trait2 - env2)^2),
        niche_prob = dnorm(env_distance, sqrt(niche_breadth)),
       
        
        # Combined probability
        combined_prob = realtive_abund*dispersal_prob * niche_prob
      )
  )) %>%
  unnest(cols = species_probs) %>%
  dplyr::select(sample, species, combined_prob) %>%
  pivot_wider(names_from = species, values_from = combined_prob) %>%
  column_to_rownames("sample")
  
  }
```

### Step 9 : Assemble local community

"either sample a fixed number of individuals or sample until a fixed number of species is reached, but both cannot be constrained simultaneously."

**fixed number of individuals**
```{r}
simulate.community.matrix<-  function(comunity_prob_matrix, num_sites){
  
  1:num_sites %>%
  purrr::map(\(x) sample(colnames(comunity_prob_matrix), size = 10000,
                  prob = comunity_prob_matrix[x,],
                  replace = T )%>%
        table %>% 
        as.data.frame %>%
        rename( sp = ".", abund = "Freq") %>%
        mutate(soil_code = paste0("sample", x))
      ) %>% 
  bind_rows() %>%
  pivot_wider(values_from = abund, names_from = sp, values_fill = 0) %>%
   column_to_rownames("soil_code")
  
  }
```



### Final function

```{r}
simulate.a.community <- function(num_sites, num_species, D, dispersal_breadth, niche_breadth, S_E_cov, phylo_tree, abund){
  # two environmental axes are independent in our two-dimensional scenario and that each only correlates with one spatial dimension.
  CVes <- matrix(c(1, 0, S_E_cov, 0,
                            0, 1, 0, S_E_cov,
                            S_E_cov, 0, 1, 0,
                            0, S_E_cov, 0, 1), nrow = 4)
  colnames(CVes) <-  c("env1", "env2", "dim1", "dim2")
  
  CVtr<- compute.CVtr(dispersal_breadth, CVes, niche_breadth, D)

  #phylo_tree <- sim.bdtree(b = 0.1, d = 0, n = num_species)  # Phylogeny of species
  
  traits_and_centroid <- simulate.characher.evolution(phylo_tree, CVtr) %>%
    #assing.abundance%>%
    mutate(abund = abund,
        realtive_abund= abund / sum(abund)) %>%
    mutate(dispersal_breadth = dispersal_breadth,
        niche_breadth = niche_breadth,
        S_E_cov = S_E_cov)
    
  spatial_and_env <- condition.and.position.samples(num_sites,CVes)
  
  create.community.matrix.probability(spatial_and_env,traits_and_centroid )%>%
    simulate.community.matrix(num_sites) -> community_matrix
  
  return(list(CVtr = CVtr,
              phylo_tree = phylo_tree,
              traits_and_centroid = traits_and_centroid,
              spatial_and_env =spatial_and_env, 
              community_matrix = community_matrix))
  }
```

### simulation

```{r}
#
num_species <- 500
num_sites <- 20
D = 2


purrr::map2(list(0.0001, 10,0.0001, 10), list(0.95, 0.95,0.3, 0.3), 
     \(x,y) simulate.a.community(num_sites = num_sites,
                                 num_species = num_species,
                                 D = D,
                                 dispersal_breadth =  x,
                                 niche_breadth =  x,
                                 S_E_cov = y)[["traits_and_centroid"]]) %>%
  bind_rows()-> obj

obj %>%
  ggplot() +
  facet_grid(cols = vars(niche_breadth), rows =vars(S_E_cov))+
  geom_point(aes(centriod2, trait2))+
  geom_smooth(aes(centriod2, trait2), method = "lm")+
  geom_abline(intercept = 0, slope = 1)+
  scale_x_continuous(limits = c(-3,3))
```

```{r}
niche_breadth <- 0.0001 # Environmental filtering strength
dispersal_breadth <-0.0001  # Dispersal limitation strength
S_E_cov = 0.95

simulate.a.community(num_sites, num_species, D, dispersal_breadth, niche_breadth, S_E_cov) -> test1


### Hill beta diversity based on abundances
test1$community_matrix %>%
  t() %>%
  hilldiv::pair_dis(qvalue=0, level="1",metric = "C") -> pairwise_beta_div

### Phylogenetic beta diversity based on abundances
phydist<-cophenetic(test1$phylo_tree)
comdist.result <- picante::comdist(test1$community_matrix, phydist,abundance.weighted = T)
mean(comdist.result)
```

```{r}
beta.div.hill <- function(mat){
  mat%>%
  t() %>%
  hilldiv::pair_dis(qvalue=0, level="1",metric = "C") -> hilldiv_list
  return(mean(hilldiv_list[["L1_beta"]], na.rm = T))
}
beta.div.phylo <- function(mat,tree){
  phylo_dist <- cophenetic(tree)
  beta_phylo <- picante::comdist(mat, phylo_dist,abundance.weighted = F)
  return(mean(beta_phylo, na.rm =T))
}
compute.div <- function(mat, tree){
  data.frame(mean_beta_hill = beta.div.hill(obj$community_matrix),
                        mean_beta_phylo = beta.div.phylo(obj$community_matrix,
                                                          obj$phylo_tree),
                        dispersal_breadth = obj$traits_and_centroid$dispersal_breadth[1] ,
                        niche_breadth  =obj$traits_and_centroid$niche_breadth[1])}
```

```{r}
num_species <- 250
num_sites <- 16
D = 2
S_E_cov = 0.7
breath_value <- round(exp(seq(log(0.0001),log(10), length.out = 10)),4)
test_fin <- data.frame()

  phylo_tree <- sim.bdtree(b = 0.1, d = 0, n = num_species)
  abund = rpois(num_species, exp(rnorm(n = num_species, 5, 2)))
  purrr::map2(as.list(rep(breath_value,10)), as.list(sort(rep(breath_value,10))), 
       \(x,y) simulate.a.community(num_sites = num_sites,
                                   num_species = num_species,
                                   D = D,
                                   dispersal_breadth =  x,
                                   niche_breadth =  y,
                                   S_E_cov = S_E_cov,
                                   phylo_tree = phylo_tree,
                                   abund = abund)) -> obj_seq_along_breaths
  
  ### Hill beta diversity based on abundances
  
  cl <- parallel::makeCluster(8)
  doParallel::registerDoParallel(cl)
  
      
  test <- foreach::foreach(obj = obj_seq_along_breaths,
                   .packages = c ("tidyverse", "hilldiv", "picante")) %dopar%{
                     1:20 %>%
                      purrr::map(\(x) simulate.a.community(num_sites = num_sites,
                                   num_species = num_species,
                                   D = D,
                                   dispersal_breadth =  x,
                                   niche_breadth =  y,
                                   S_E_cov = S_E_cov,
                                   phylo_tree = obj$phylo_tree,
                                   abund = sample(abund))) %>%
                       compute.div(phylo_tree)
                   }
    
    
     data.frame(mean_beta_hill = beta.div.hill(obj$community_matrix),
                        mean_beta_phylo = beta.div.phylo(obj$community_matrix,
                                                          obj$phylo_tree),
                        dispersal_breadth = obj$traits_and_centroid$dispersal_breadth[1] ,
                        niche_breadth  =obj$traits_and_centroid$niche_breadth[1])
  
  parallel::stopCluster(cl)
     
  test %>%
    bind_rows() -> test
  test$n_sim <- n_sim
  test_fin <- rbind(test_fin,test)


test_fin

test_fin %>%
  group_by(n_sim, dispersal_breadth, niche_breadth) %>%
  summarise(mean_beta_hill = mean(mean_beta_hill),
            mean_beta_phylo = mean(mean_beta_phylo ))%>%
  mutate(mean_beta_hill = (mean_beta_hill - mean(mean_beta_hill))/sd(mean_beta_hill),
         mean_beta_phylo  = (mean_beta_phylo - mean(mean_beta_phylo))/sd(mean_beta_phylo) )%>%
  pivot_longer(starts_with("mean"))%>%
  ggplot() +
  facet_wrap(~name, scales = "free") +
  geom_raster(aes(dispersal_breadth,niche_breadth, fill = value))+
  scale_x_log10()+
  scale_y_log10()

```


```{r}
library(sbm)

community_matrix_sim%>%
  column_to_rownames("soil_code") %>%
  as.matrix() %>%

  log1p %>%
  plotMyMatrix(dimLabels = c("soil_code", "species"), plotOptions = list(rowNames = T,colNames = T))+
  theme(
    legend.position="none",
    axis.text.x =element_text(angle = 45, vjust = 1, hjust = 1, size = 6, face="italic"),
    axis.text.y = element_text(size = 5,face="italic"))+
  theme(strip.text.x = element_text(colour = "gray", size=15, face ="italic"),
        strip.text.y  = element_text(colour = "gray", size=15, face ="italic"),
        strip.background = element_rect(fill="white"),
        panel.border = element_rect(color = "black",  fill = NA))
```

### step 10 : compute beta diversities

```{r}


### Hill beta diversity based on abundances
community_matrix_sim %>%
  t() %>%
  hilldiv::pair_dis(qvalue=0, level="1",metric = "C") -> pairwise_beta_div

### Phylogenetic beta diversity based on abundances
phydist<-cophenetic(phylo_tree)
comdist.result <- picante::comdist(community_matrix_sim, phydist,abundance.weighted = T)
```

### Some visualisation

```{r}
spatial_env %>%
  ggplot() +
  geom_point(aes(dim1, dim2, col = env1), cex =3, alpha = 0.9)
spatial_env %>%
  ggplot() +
  geom_point(aes(dim1, dim2, col = env2), cex =3, alpha = 0.9)
traits_and_centroid%>%
  ggplot() +
  geom_point(aes(centriod1, centriod2, col = trait1), cex =3, alpha = 0.9)
```


```{r}
pairwise_beta_div[[1]]%>%
  as.data.frame()%>%
  rownames_to_column("soil_code")%>%
  pivot_longer(-soil_code, names_to = "soil_code2", values_to = "obs_beta_div") %>%
  filter(!is.na(obs_beta_div))  -> pairwise_beta_div_df

comdist.result%>%
  as.data.frame()%>%
  rownames_to_column("soil_code")%>%
  pivot_longer(-soil_code, names_to = "soil_code2", values_to = "obs_beta_phylo") %>%
  filter(!is.na(obs_beta_phylo))  -> pairwise_beta_phylo_df


spatial_env %>%
  as.data.frame() %>%
  rownames_to_column("soil_code")%>%
  dplyr::mutate(soil_code =paste("sample",soil_code, sep = "")) -> spatial_env


pairwise_beta_div_df %>%
  left_join(spatial_env, by = "soil_code") %>%
  left_join(spatial_env, by = join_by(soil_code2 == soil_code)) %>%
  mutate(diff_env1 = abs(env1.x-env1.y),
         diff_env2 = abs(env2.x-env2.y),
         diff_dim1 = abs(dim1.x-dim1.y),
         diff_dim2 = abs(dim2.x-dim2.y)) %>%
  dplyr::select(-ends_with("x"),-ends_with("y"))-> beta_div_spatial_env_df


beta_div_spatial_env_df %>%
  pivot_longer(starts_with("diff"), values_to = "diff_val", names_to = "diff_var")%>%
  ggplot()+
  facet_wrap(~diff_var)+
  geom_point(aes(diff_val,obs_beta_div )) +
  labs(x = "differences saptial or envrionnemental", y = "beta diversity") +
  scale_y_continuous(limits = c(1,2))+
  main_theme


```

