---
title: "Untitled"
author: "Mathis Gheno"
date: "2024-10-03"
output: html_document
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
title: "01_metho_comparision_taxo"
author: "Mathis Gheno"
date: "2024-09-20"
output:
  html_document:
    df_print: paged
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

## load required packages

```{r packages, message=FALSE}
library(tidyverse)
library(ggforce)
library(ade4) # Coinertia
library(vegan) # Rarefaction
library(ggrepel)
library(metacoder)
library(phyloseq)
library(hilldiv)

library(rjags)
library(ggmcmc)
library(coda)

library(parallel)
library(doParallel)
library(foreach)

main_theme = theme_minimal()+
  theme(line = element_blank(), 
        axis.line = element_line(colour = "black"),
        panel.border = element_blank(),
        axis.ticks =  element_line(colour = "black"),
        axis.text.x = element_text(colour = "black", size=22, face="italic", angle = 45, vjust = 1, hjust = 1),
        axis.text.y = element_text(colour = "black", size=22, face="italic"),
        legend.title = element_text(colour = "black", size=20,
                                    hjust =0.5),
        
        legend.text = element_text(colour = "black", size=18),
        axis.title= element_text(size=28),
        strip.text = element_text(colour = "black", size=15, face ="italic"))
```

## Data path

```{r}
metab1 = "data/metabarcoding_vs_metagenomics/oracle_metaBt0_16S_samples_run_20190715_kraken2_assignment_genuses.tsv"
metab2 = "data/metabarcoding_vs_metagenomics/oracle_metaBt0_16S_samples_run_20200106_kraken2_assignment_genuses.tsv"
metag1 = "data/metabarcoding_vs_metagenomics/oracle_metaGt0_SAMA_12_samples_kraken2_SSU_assignment_genuses.tsv"
metag2 = "data/metabarcoding_vs_metagenomics/oracle_metaGt0_SAMA_21_1_samples_kraken2_SSU_assignment_genuses.tsv"
metag3 = "data/metabarcoding_vs_metagenomics/oracle_metaGt0_SAMA_21_2_samples_kraken2_SSU_assignment_genuses.tsv"
```

# Load data

## Set up function to load and pepare data

```{r}
taxonomic_levels <- c("kingdom", "phylum", "class", "order")# "family", "genus")
## ---- Arrange metaB ----

# Select data from bracken method, and rename columns 
select.and.rename.cols = function(tab, method = "B"){
  tab %>%
    read_tsv(col_names = T, skip = 1) %>%
    select(contains("bracken_genuses"), `#OTU ID`, taxonomy) %>%
    rename_with( # rename colnames
      . %>%
      str_remove( "_bracken_genuses") %>% 
      str_replace_all("^OR-", "BU_") %>% 
      str_replace_all("-", "_") %>%
      str_remove("_S\\d+") %>%
      #  (^T_) Select strings with the letter T at the beginning of the string; (\\d) detect number ; 
      #  "CONT_BU_PCR_\\2" replace name to matching string and add group 2 (which correspond to  (\\d) )
      str_replace("(^T_)(\\d)", "CONT_BU_PCR_\\2") %>% 
      str_replace("(BU_T_extr)", "CONT_BU_ext") %>%
      str_replace("(^T_Ex)", "CONT_BU_ext") %>%
      str_replace("#OTU ID", "OTU") %>%
      # add a 0 in matching sting eg. BU_SER_1_R1 -> BU_SER_01_R1
      str_replace("(BU_[A-Z]+_)(\\d{1})(_R\\d+)", "\\10\\2\\3")
      ) %>%
    rename_with(~ paste0(., "_", method), contains("BU")) %>% # add B or G at the end of soil code depending on the method (metabarcoding or metagenomic)
    mutate(taxonomy = str_remove(taxonomy, "; [f]__.+")) %>%
    separate_wider_delim(taxonomy, names = taxonomic_levels, delim = "; ", cols_remove = F) # Separate taxonomy column in columns of taxonomic levels
    # filter(kingdom == "k__Bacteria")
}

```

## Decontamination function

Compute the total number of reads for each OTU present in control samples. That sum is then substracted from all occurrences of that OTU in true samples. The rationale is as follows:

-   if an OTU is abundant in control samples, but rare in true samples, then it is a contamination specific to the control samples, and it will be eliminated by the substraction (i.e, final abundance is 0),
-   if an OTU is present in control samples, and present in true samples (systematic contamination, will be mitigated by the substraction),
-   if an OTU is rare in control samples, but abundant in true samples (cross-talk, will be eliminated/mitigated by the substraction)

Control samples can be eliminated from the statistical analysis after the substraction (all OTUs present in control samples have been zeroed out).

```{r}
. %>%
  replace(. == 0, NA) %>%
  pivot_longer(starts_with("BU"), names_to = "samples", values_to = "reads") %>%
  filter(!is.na(reads)) %>%
  # merge control samples
  mutate( n = rowSums(across(starts_with("CONT")), na.rm = T)
          ) %>%
  # substract abundance of control samples
  mutate(reads = case_when(
    is.na(n)  ~ reads,
    n > reads ~ 0,
    TRUE      ~ reads - n)
    ) %>%
  select(-n,-starts_with("CONT")) %>%
  pivot_wider(values_from = reads, names_from = samples, values_fill = 0) -> decontaminate


```

## Load and format data of both method in the same dataframe

```{r}
metab1 %>%
 select.and.rename.cols("B") %>%
 decontaminate -> metab1_decont_table

metab2 %>%
 select.and.rename.cols("B") %>%
 decontaminate -> metab2_decont_table

# Merge run 1 and 2 of metaB, because some sample failed in the first run
metab1_decont_table %>%
  full_join(metab2_decont_table, by = join_by(OTU == OTU, kingdom == kingdom, phylum == phylum,
                                               class == class, order == order, taxonomy == taxonomy)
            #family == family,#genus == genus
            ) %>%
  select(-ends_with(".x")) %>% # Remove conflicting columns from the left-hand side
  rename_with(~ sub(".y", "", .), ends_with(".y")) %>%  # Rename right-hand columns
  mutate(across(where(is.numeric), 
                ~ replace(., is.na(.), 0))
         ) -> metab_decont_table

          
metag1 %>%
 select.and.rename.cols("G") %>%
 decontaminate -> metag1_decont_table

metag1_decont_table %>%
  full_join(metab_decont_table, by = join_by(OTU == OTU, kingdom == kingdom, phylum == phylum,
                                               class == class, order == order, taxonomy == taxonomy)
            # family == family, genus == genus
            ) %>%
   mutate(across(where(is.numeric),
                 ~ replace(., is.na(.), 0))
          ) %>%
   select(-c(BU_TS_15_R1_B, BU_TS_15_R2_B)) -> meta_g_b_decond_table # remove sample because missing for metaG

```
## Diversity analysis

### Rarefaction function

```{r}
rarefaction.func <- function(df,rar_sample, n_rare = 100, rarcur = T){
  
  
  if(rarcur){ # {{ if rarefaction curve needed}}
    metab1_decont_table %>%
      select(starts_with("BU")) %>%
      t() %>%
      vegan::rarecurve(step = 20, sample = rar_sample, col = "blue", cex = 0.6)
  }
   df %>%
    select(starts_with("BU")) %>%
    dim -> D
   # {{rarefaction}}
  df_rarefy = matrix(0, nrow =  D[1], ncol = D[2])
  for(n in 1:n_rare){
  df %>%
    select(starts_with("BU")) %>%
    t() %>%
    vegan::rrarefy(rar_sample) %>%
    t() + df_rarefy -> df_rarefy
    
  }
  round(df_rarefy/n_rare) -> df_rarefy
  df_rarefy %>%
    bind_cols(
      df %>%
        select(!starts_with("BU"))
    ) -> df_rarefy
  return(df_rarefy)
}

```

### Sequencing depth per method and per samples (Libraries sizes)

```{r}
meta_g_b_decond_table %>%
  pivot_longer(starts_with("BU"), names_to = "sample_ID", values_to = "reads") %>%
  group_by(sample_ID) %>%
  summarise(reads_sum = sum(reads)) %>% 
  arrange(reads_sum) -> lib_size

lib_size$reads_sum[1] -> min_ech

lib_size %>%
  mutate(method = str_extract(sample_ID, "R\\d_[A-Z+]")) %>%
  ggplot()+
  geom_point(aes(x = reorder(sample_ID, reads_sum), y = reads_sum, col = method))+
  scale_y_log10(limits = c(0.01, 10^6))+
  labs(x = "Sample ID", y = "Reads sum per sample (log10)", col ="Method")+
  theme_classic()+
  theme(axis.text.x = element_text(colour = "black", size=5, face="italic", angle = 45, vjust = 1, hjust = 1),
        axis.text.y = element_text(colour = "black", size=22, face="italic"),
        legend.title = element_text(colour = "black", size=20,
                                    hjust =0.5),
        
        legend.text = element_text(colour = "black", size=18))
```

### Rarefaction

```{r}
set.seed(1000)
meta_g_b_decond_table %>%
  rarefaction.func(rar_sample = min_ech, rarcur = T) -> meta_g_b_decond_table_rare

meta_g_b_decond_table_rare %>% 
  select(starts_with("BU")) %>% colSums

```

### Alpha Diversity comparison

```{r}
#  richness
dR <- meta_g_b_decond_table_rare %>% 
  column_to_rownames("OTU") %>%
  select(starts_with("BU")) %>%
  hilldiv::hill_div(qvalue = 0) %>%
  as_tibble(rownames = NA) %>%
  rownames_to_column(var = "samples") %>%
  mutate(method = str_extract(samples, "R\\d_[BG]"),
         type = "richness")
  

#  shannon diversity as hill numbers; rare species
dREr <- meta_g_b_decond_table_rare %>% 
  column_to_rownames("OTU") %>%
  select(starts_with("BU")) %>%
  hilldiv::hill_div(qvalue = 1) %>%
  as_tibble(rownames = NA) %>%
  rownames_to_column(var = "samples") %>%
  mutate(method = str_extract(samples, "R\\d_[BG]"),
         type = "shannon")

## inverse Simpson = simpson diversity as hill numbers; abundant species
dREa <-meta_g_b_decond_table_rare %>% 
  column_to_rownames("OTU") %>%
  select(starts_with("BU")) %>%
  hilldiv::hill_div(qvalue = 2) %>%
  as_tibble(rownames = NA) %>%
  rownames_to_column(var = "samples") %>%
  mutate(method = str_extract(samples, "R\\d_[BG]"),
         type = "simpson")

# evenness (= shannon evenness) as hill ratio
eDRr <- dREr %>%
  left_join(dR, by = "samples") %>%
  mutate(value = value.x / value.y) %>%
  select(samples,value) %>%
  mutate(method = str_extract(samples, "R\\d_[BG]"),
         type = "eveness_sha")

#{{evenness (= ssimpson evenness) as hill ratio}}
eDRa  <- dREa %>%
  left_join(dR, by = "samples") %>%
  mutate(value = value.x / value.y) %>%
  select(samples,value) %>%
  mutate(method = str_extract(samples, "R\\d_[BG]"),
         type = "eveness_sim")

df_diversites <- dR %>%
  bind_rows(dREr,dREa,eDRr,eDRa) %>%
  mutate(soil_code = str_extract(samples, "BU_[A-Z]+_\\d+")) %>%
  select(-samples) %>%
  pivot_wider(names_from = method, values_from = value )

rm(dREr,dREa,eDRr,eDRa)
```

```{r}
df_diversites%>%
  ggplot()+
  geom_abline(slope = 1, col = "red")+
  geom_point(aes(R1_B,R2_B))+
  geom_smooth(aes(R1_B,R2_B), method = "lm")+
  facet_wrap(~type, scale = "free")+
  main_theme

df_diversites%>%
  ggplot()+
  geom_abline(slope = 1, col = "red")+
  geom_point(aes(R1_G,R2_G))+
  geom_smooth(aes(R1_G,R2_G), method = "lm")+
  facet_wrap(~type, scale = "free")+
  main_theme

df_diversites%>%
  ggplot()+
  geom_abline(slope = 1, col = "red")+
  geom_point(aes(R1_B,R1_G))+
  geom_smooth(aes(R1_B,R1_G), method = "lm")+
  facet_wrap(~type, scale = "free")+
  main_theme

df_diversites%>%
  ggplot()+
  geom_abline(slope = 1, col = "red")+
  geom_point(aes(R2_B,R2_G))+
  geom_smooth(aes(R2_B,R2_G), method = "lm")+
  facet_wrap(~type, scale = "free")+
  main_theme
```

- Diversities are really similar between R1 and R2 of metabarcoding. Both strand approximate the same number of taxa and same Eveness.
- Richness is a bit higher in R2 of metabargenomic compaire to R1 (R2 detetect approximately 20 more taxa than R1). 
There is a small positive correlation of evenness between the 2 stand (Surprising !)
- There is almost no correlation of richness, diversities and evneness for comparisons between metabarcoding and metagenomics

```{r}
meta_g_b_decond_table_rare %>% select(starts_with("BU")) -> temp
plot(1:length(colSums (temp!=0 )), colSums (temp!=0 ))
```


### Beta diversities

```{r}
meta_g_b_decond_table_rare %>%
  select(starts_with("BU")) %>%
  colnames() %>%
  as.data.frame() %>%
  rename(sample = ".") %>%
  mutate(groups = str_extract(sample, "R\\d_[BG]")) -> hierarchie_meta_g_b
  
meta_g_b_decond_table_rare %>%
  select(starts_with("BU")) %>% 
  hilldiv::pair_dis(qvalue=0, level = "1") -> beta_div

beta_div[[1]]%>%
  as.data.frame()%>%
  rownames_to_column("sample")%>%
  pivot_longer(-sample) %>%
  filter(!is.na(value)) %>%
  mutate(group1 = str_extract(sample, "R\\d_[BG]"),
         group2 = str_extract(name, "R\\d_[BG]"),
         soil_code1 = str_extract(sample, "BU_[A-Z]+_\\d+"),
         soil_code2 = str_extract(name, "BU_[A-Z]+_\\d+"),
         group_comp = paste(group1,"x",group2)) %>%
  filter(soil_code1 == soil_code2) %>%
  ggplot()+
  #facet_wrap(~group2 * group1)+
  geom_sina(aes(group_comp, value),maxwidth = 1, alpha = 0.5, cex = 5, jitter_y = F) +
  labs(x = "compared method", y = "beta diversity") +
  scale_y_continuous(limits = c(1,2))+
  main_theme


```


```{r}
meta_g_b_decond_table_rare %>%
  select(starts_with("BU"), OTU) %>% 
  pivot_longer(-OTU)%>%
  mutate(name = str_remove(name, "_R\\d"))%>%
  group_by(name, OTU)%>%
  summarise(value = sum(value))%>%
  pivot_wider(names_from = name, values_from = value) %>% select(starts_with("BU")) %>%
  hilldiv::pair_dis(qvalue=0, level = "1") -> beta_div_merge


beta_div_merge[[1]]%>%
  as.data.frame()%>%
  rownames_to_column("sample")%>%
  pivot_longer(-sample) %>%
  filter(!is.na(value)) %>%
  mutate(soil_code1 = str_extract(sample, "BU_[A-Z]+_\\d+"),
         soil_code2 = str_extract(name, "BU_[A-Z]+_\\d+")) %>%
  filter(soil_code1 == soil_code2) %>%
  ggplot()+
  #facet_wrap(~group2 * group1)+
  geom_sina(aes("G vs B", value),maxwidth = 0.4, alpha = 0.5, cex = 5, jitter_y = F) +
  labs(x = "compared method", y = "beta diversity")+
  main_theme

```

## Variance and mean analisys

### Variance and mean distribution of taxon (genus) per method 

```{r}

meta_g_b_decond_table_rare %>% 
  filter(rowSums(across(starts_with("BU")), na.rm = T) != 0) %>%
  pivot_longer(starts_with("BU"), names_to = "sample_ID", values_to = "reads") %>%
  mutate(method = str_extract(sample_ID, "R\\d_[A-Z+]")) %>%
  group_by(OTU, method) %>%
  summarise(reads_var = var(reads, na.rm = T),
            reads_mean = mean(reads, na.rm = T)) %>%
  ungroup() -> summarise_df

summarise_df %>%
  ggplot()+
  facet_wrap(~method)+
  geom_histogram(aes(x = reads_mean))+
  scale_x_log10()

summarise_df %>%
  ggplot()+
  facet_wrap(~method)+
  geom_histogram(aes(x = reads_var))+
  scale_x_log10()

```

Metagenomic miss more OTU that metabarcoding. Without accounting OTU at 0 abundances, mean and variance distribution of OTU number of reads look like similar for both method. Now I will check the relation between variance and mean to see if data follow a poisson distribution without zero. (In poisson distib mean = variance)

### Relation between variance and mean 

```{r}
mod_mean_v_var <- lm(log10(reads_var)~log10(reads_mean)*method,
                     data = summarise_df%>%
                              filter(reads_mean>1, reads_var>1)
                     )
summary(mod_mean_v_var)
temp = data.frame(method = c("R1_B", "R2_B", "R1_G", "R2_G"), 
                  slope = c(coef(mod_mean_v_var)[2],coef(mod_mean_v_var)[2] , coef(mod_mean_v_var)[2]+ coef(mod_mean_v_var)[6], coef(mod_mean_v_var)[2] + coef(mod_mean_v_var)[8]),
                  intercept =  c(coef(mod_mean_v_var)[1], coef(mod_mean_v_var)[1], coef(mod_mean_v_var)[1], coef(mod_mean_v_var)[1] + coef(mod_mean_v_var)[5]))
summarise_df%>%
  left_join(temp, by = "method")%>%
  filter(reads_mean>1, reads_var>1)%>%
  ggplot()+
  facet_wrap(~method)+
  geom_point(aes(log10(reads_mean), log10(reads_var)))+
  geom_smooth(aes(log10(reads_mean), log10(reads_var)), linewidth= 1.5, method = "lm")+
  geom_abline(aes(intercept = intercept, slope =1), col = "red", size = 1.5, linetype = 2)+
  geom_abline(aes(intercept = intercept, slope =2), col = "purple", size = 1.5, linetype = 2)+
  scale_x_continuous(limits = c(-2, 7))+
  scale_y_continuous(limits = c(-2, 7))

temp


```

The slopes are around 1 for R1_G and R2_G. For metagenomic species means are proportionnal to there variance ($\mu \prop \sigma^2$). For metabarcoding slopes are 1.40 for R1_B and R2_B, this tells us that for this method variances are propotional to a power law of means ($\mu^n \propto \sigma^2$ with $n = 1.40$ here)

### Simulation of 2 prossec leading to different relation between variance and mean 

```{r}
true_com <- c(rep("A",1000), rep("B",100), rep("C",10))

### Additive 
n_sim = 1000
count_df = data.frame(A = rep(NA,n_sim), B = rep(NA,n_sim), C = rep(NA,n_sim))
prop_df = data.frame(A = rep(NA,n_sim), B = rep(NA,n_sim), C = rep(NA,n_sim))

# Sampling function
for(n in 1:n_sim){
  ech =  sample(true_com, 200, replace = F)
  n_A = sum(ech == "A")
  n_B = sum(ech == "B")
  n_C = sum(ech == "C")
  count_df[n,] =c(n_A, n_B, n_C)
  prop_df[n,] =c(n_A, n_B, n_C)/200
  
}
count_df%>%
  pivot_longer(everything(), names_to = "sp", values_to = "abund") %>%
  group_by(sp) %>%
  summarise(
    var_sp = var(abund),
    mean_sp = mean(abund),
    sd_sp = sd(abund)
  ) -> df_summar

ggplot(df_summar)+
  geom_point(aes(mean_sp, var_sp))+
  geom_abline(intercept = 0, slope = 1, col = "red", size = 1.5, linetype = 2)+
  geom_abline(intercept = 0, slope = 2, col = "purple", size = 1.5, linetype = 2)+
  scale_x_log10(limits = c(1,1000))+
  scale_y_log10(limits = c(1,1000))

### Multiplicative 
amplifi_df = data.frame(A = rep(NA,n_sim), B = rep(NA,n_sim), C = rep(NA,n_sim))
depth = 10
# Amplification function
for(n in 1:n_sim){
  val = c(1000,100,10)
  for(d in 1:depth){
    temp <- runif(3,1,2)
    val = val*temp
  }
  true_com <- c(rep("A",val[1]), rep("B",val[2]), rep("C",val[3]))
  ech =  sample(true_com, 200, replace = F)
  n_A = sum(ech == "A")
  n_B = sum(ech == "B")
  n_C = sum(ech == "C")
  amplifi_df[n,] =c(n_A, n_B, n_C)
}

amplifi_df%>%
  pivot_longer(everything(), names_to = "sp", values_to = "abund") %>%
  group_by(sp) %>%
  summarise(
    var_sp = var(abund),
    mean_sp = mean(abund),
    sd_sp = sd(abund)
  ) -> df_amplifi_summar

ggplot(df_amplifi_summar)+
  geom_point(aes(mean_sp, var_sp))+
  geom_abline(intercept = 0, slope = 1, col = "red", size = 1.5, linetype = 2)+
  geom_abline(intercept = 0, slope = 2, col = "purple", size = 1.5, linetype = 2)+
  scale_x_log10(limits = c(1,1000))+
  scale_y_log10(limits = c(1,1000))
```





## Taxa between method comparision analasys (Metacoder and phyloseq)

Function that compute differential gene analysis with edgeR method from a phyloseq object

```{r}
phyloseq_to_edgeR = function(physeq, group, method="RLE", ...){
  require("edgeR")
  require("phyloseq")
  # Enforce orientation.
  if(!taxa_are_rows(physeq) ){ physeq <- t(physeq)}
  x = as(otu_table(physeq), "matrix")
  # Add one to protect against overflow, log(0) issues.
  x = x + 1
  # Check `group` argument
  if( identical(all.equal(length(group), 1), TRUE) & nsamples(physeq) > 1 ){
    # Assume that group was a sample variable name (must be categorical)
    group = get_variable(physeq, group)
  }
  # Define gene annotations (`genes`) as tax_table
  taxonomy = tax_table(physeq, errorIfNULL=FALSE)
  if( !is.null(taxonomy) ){
    taxonomy = data.frame(as(taxonomy, "matrix"))
  } 
  # Now turn into a DGEList
  y = DGEList(counts=x, group=group, genes=taxonomy, remove.zeros = TRUE, ...)
  # Calculate the normalization factors
  z = calcNormFactors(y, method=method)
  # Check for division by zero inside `calcNormFactors`
  if( !all(is.finite(z$samples$norm.factors)) ){
    stop("Something wrong with edgeR::calcNormFactors on this data,
         non-finite $norm.factors, consider changing `method` argument")
  }
  # Estimate dispersions
  return(estimateTagwiseDisp(estimateCommonDisp(z)))
}
```

Function that compute differential analysis and return an object compatible to metacoder,
to plot it on the taxonomic comparison tree

```{r}
differential.analysis <- function(obj, pair){
  
  # Extract taxon rank associated to each Metacoder taxon_id
  rank_df = data.frame(taxon_rank = sapply(names(obj$taxa), function(x) obj$taxa[[x]]$rank$name)) %>%
  rownames_to_column("taxon_id")
  
  # Join taxon rank and taxon total abundances data frame
  rank_df %>%
    left_join(obj$data$tax_abund, by = "taxon_id") %>%
  tibble -> tax_abund
  
  # Create sample data (data frame that indicate sample's condition/method belonging)
  data.frame(sample_ID = colnames(tax_abund %>%
                                    select(starts_with("BU")))) %>%
      mutate(method = str_extract(sample_ID, "R\\d_[A-Z+]$")) %>%
      column_to_rownames("sample_ID") -> sampledata
  
  # Initialize output dataframe
  pair_wise_taxa_df = data.frame()
  t_rank = "c"
  for(t_rank in unique(tax_abund$taxon_rank)){
    
    # Compute OTU table with sample from both method in colums and taxa in rows
    tax_abund %>%
      filter(taxon_rank == t_rank) %>%
      select(starts_with("BU"), taxon_id) %>%
      column_to_rownames("taxon_id") -> otu_tab
    
    # Compute taxon table, taxon_id as row name and in a column. taxon_id as row needed  
    # to create phyloseq object. And I need to have a column that discriminates taxon, which taxon_id from 
    # meta coder do perfectly.(I could recover the full taxa description (Kindom, class, ...),
    # but it is not useful because taxon_id is analogue.)
    tax_abund %>%
      filter(taxon_rank == t_rank) %>%
      select(taxon_id )%>%
      mutate(taxon_id2 = taxon_id) %>%
      column_to_rownames("taxon_id2") %>%
      as.matrix -> tax_tab
  
    physeq_obj = phyloseq(otu_table(otu_tab, taxa_are_rows = T),
                          tax_table(tax_tab),
                          sample_data(sampledata))
  
    dge = phyloseq_to_edgeR(physeq_obj, group="method")
    et = exactTest(dge, pair=c(pair[2], pair[1]))
    # Extract values from test results
    tt = topTags(et, n=nrow(dge$table), adjust.method="fdr", sort.by="PValue")
    res = tt@.Data[[1]]
    
    pair_wise_taxa_df %>%
      bind_rows(res %>%
                  select(taxon_id, logFC, logCPM, PValue, FDR))%>%
      tibble() -> pair_wise_taxa_df
  }
  return(pair_wise_taxa_df)
}
```

Function that make the metacoder object that contain the taxonomic comparison tree, rank comparison (Mann-Whitney/Wilcoxon test)

```{r}
make.tax.tree.comp = function(df, pair){
  df %>%
  select(OTU, kingdom, phylum, class, order, taxonomy, ends_with(pair[1]), ends_with(pair[2])) %>% # family, genus,
  # create metacoder object
  metacoder::parse_tax_data(class_cols = "taxonomy", # The column in the input table
                      class_sep = "; ",
                      class_regex = "^([a-z]{0,1})_{0,2}(.*)$",
                      class_key = c("tax_rank" = "taxon_rank", "name" = "taxon_name")) -> obj
  
  ##### compute abudance per taxon for the 2 paire
  obj$data$tax_abund <- metacoder::calc_taxon_abund(obj, "tax_data",
                                       cols = startsWith(colnames(obj$data$tax_data),"BU"))
  
  obj$data$tax_abund_group <- metacoder::calc_taxon_abund(obj, "tax_data",
                                       cols = startsWith(colnames(obj$data$tax_data),"BU"),
                                                         groups = str_extract(str_subset(colnames(obj$data$tax_data), "_R\\d_[A-Z+]"), "R\\d_[A-Z+]"))
  obj$data$diff_table <- metacoder::compare_groups(obj, "tax_abund",
                                       cols = startsWith(colnames(obj$data$tax_abund),"BU"),
                                                         groups = str_extract(str_subset(colnames(obj$data$tax_abund), "_R\\d_[A-Z+]"), "R\\d_[A-Z+]"))
  obj <- mutate_obs(obj, "diff_table",
                  wilcox_p_value = p.adjust(wilcox_p_value, method = "fdr"))
  # Ns. set to 0
  obj$data$diff_table %>%
    mutate(median_diff = case_when(wilcox_p_value > 0.05 ~ 0,
                      .default  = median_diff),
           log2_median_ratio = case_when(wilcox_p_value > 0.05 ~ 0,
                      .default  = median_diff)) -> obj$data$diff_table
  
  ##### differential analysis of sequence read count data with edgeR
  obj$data$diff_analysis_table <- differential.analysis(obj = obj, pair = pair)
  
  # Ns. set to 0
  obj$data$diff_analysis_table %>%
    mutate(logFC = case_when(FDR > 0.05 ~ 0,
                      .default  = logFC),
           logCPM = case_when(FDR > 0.05 ~ 0,
                      .default  = logCPM)) -> obj$data$diff_analysis_table
  
  return(obj)
}

```

### Metabarcoding R1 vs R2

```{r}
pal = colorspace::diverging_hcl(3, palette = "Tropic")
meta_g_b_decond_table %>%
  make.tax.tree.comp(pair = c("R1_B","R2_B")) -> obj1

meta_g_b_decond_table_rare %>%
  make.tax.tree.comp(pair = c("R1_B","R2_B")) -> obj1_rare
```

####  Wilcox 2 by 2 comparison

```{r}
obj1 %>%
  metacoder::filter_taxa(taxon_ranks == "o", supertaxa = TRUE)%>%
  metacoder::heat_tree(node_label = taxon_names,
          node_size = n_obs, # number of OTUs
          node_color = log2_median_ratio, # difference between groups
          node_color_interval = c(-max(abs(log2_median_ratio)), max(abs(log2_median_ratio))), # symmetric interval
          node_color_range = pal, # diverging colors
          node_size_axis_label = "OTU count",
          node_color_axis_label = "Log 2 ratio of median counts",
          layout = "davidson-harel", initial_layout = "reingold-tilford",
          title = paste(unique(treatment_1) ,"vs", unique(treatment_2),"(wilcox)"))

obj1_rare %>%
  metacoder::filter_taxa(taxon_ranks == "o", supertaxa = TRUE)%>%
  metacoder::heat_tree(node_label = taxon_names,
          node_size = n_obs, # number of OTUs
          node_color = log2_median_ratio, # difference between groups
          node_color_interval = c(-max(abs(log2_median_ratio)), max(abs(log2_median_ratio))), # symmetric interval
          node_color_range = pal, # diverging colors
          node_size_axis_label = "OTU count",
          node_color_axis_label = "Log 2 ratio of median counts",
          layout = "davidson-harel", initial_layout = "reingold-tilford",
          title = paste(unique(treatment_1) ,"vs", unique(treatment_2), "(wilcox/rare)"))
```


```{r}
sapply(obj1$taxa, function(x) as.character(x$get_name())) %>% 
  as.data.frame %>%
  rename(taxa = ".")%>%
  rownames_to_column("taxon_id") -> taxo1_df

sapply(obj1_rare$taxa, function(x)as.character(x$get_name())) %>% 
  as.data.frame %>%
  rename(taxa = ".") %>%
  rownames_to_column("taxon_id")-> taxo2_df

table(taxo1_df == taxo2_df)# must be only TRUE

obj1$data$diff_table %>%
  select(taxon_id, log2_median_ratio) %>%
  left_join(obj1_rare$data$diff_table %>%
            select(taxon_id, log2_median_ratio), by ="taxon_id" ) %>% 
  ggplot()+
  geom_point(aes(log2_median_ratio.x/5, log2_median_ratio.y),size = 3, alpha =0.6)+
  geom_abline(intercept = 0, slope = 1, size = 2, col = "red")+
  labs(x =" log2(m1/m2)/5 No rarefaction", y ="log2 (m1/m2) rarefaction")+
  main_theme
```



#### Differential expression of replicated count data

```{r}
obj1 %>%
  metacoder::filter_taxa(taxon_ranks == "o", supertaxa = TRUE) %>%
  metacoder::heat_tree(node_label = taxon_names,
          node_size = n_obs, # number of OTUs
          node_color = logFC, # difference between groups
          node_color_interval = c(-max(abs(logFC)), max(abs(logFC))), # symmetric interval
          node_color_range = pal, # diverging colors
          node_size_axis_label = "OTU count",
          node_color_axis_label = "logFC",
          layout = "davidson-harel", initial_layout = "reingold-tilford",
          title = paste(unique(treatment_1) ,"vs", unique(treatment_2),"DGE"))

obj1_rare %>%
  metacoder::filter_taxa(taxon_ranks == "o", supertaxa = TRUE) %>%
  metacoder::heat_tree(node_label = taxon_names,
          node_size = n_obs, # number of OTUs
          node_color = logFC, # difference between groups
          node_color_interval = c(-max(abs(logFC)), max(abs(logFC))), # symmetric interval
          node_color_range = pal, # diverging colors
          node_size_axis_label = "OTU count",
          node_color_axis_label = "logFC",
          layout = "davidson-harel", initial_layout = "reingold-tilford",
          title = paste(unique(treatment_1) ,"vs", unique(treatment_2),"DGE rare"))
```


```{r}
sapply(obj1$taxa, function(x) as.character(x$get_name())) %>% 
  as.data.frame %>%
  rename(taxa = ".") %>%
  rownames_to_column("taxon_id") -> taxo1_df_DGE

sapply(obj1_rare$taxa, function(x)as.character(x$get_name())) %>% 
  as.data.frame %>%
  rename(taxa = ".") %>%
  rownames_to_column("taxon_id")-> taxo2_df_DGE

obj1$data$diff_analysis_table %>%
  select(taxon_id, logFC) %>%
  left_join(obj1_rare$data$diff_analysis_table %>%
            select(taxon_id, logFC), by ="taxon_id" ) %>% 
  ggplot()+
  geom_point(aes(logFC.x, logFC.y),size = 3, alpha =0.6)+
  geom_abline(intercept = 0, slope = 1, size = 2, col = "red")+
  labs(x =" log2(m1/m2) No rarefaction", y ="log2 (m1/m2) rarefaction")+
  main_theme
  
```

### Metagenomic R1 vs R2

```{r}
meta_g_b_decond_table %>%
  make.tax.tree.comp(pair = c("R1_G","R2_G")) -> obj2

meta_g_b_decond_table_rare %>%
  make.tax.tree.comp(pair = c("R1_G","R2_G")) -> obj2_rare
```

Different pattern for metaG than metaB here in phylum *Protobacteia* the class *Gammaprotobacteria* and *Alphaprotobacteria* are both more detected by R1. *Fimicutes* and *Actinobacteria* are also better detected by R1

####  Wilcox 2 by 2 comparison

```{r}
obj2 %>%
  metacoder::filter_taxa(taxon_ranks == "o", supertaxa = TRUE) %>%
  metacoder::heat_tree(node_label = taxon_names,
          node_size = n_obs, # number of OTUs
          node_color = log2_median_ratio, # difference between groups
          node_color_interval =c(-max(abs(log2_median_ratio)), max(abs(log2_median_ratio))), # symmetric interval
          node_color_range = pal, # diverging colors
          node_size_axis_label = "OTU count",
          node_color_axis_label = "Log 2 ratio of median counts",
          layout = "davidson-harel", initial_layout = "reingold-tilford",
          title = paste(unique(treatment_1) ,"vs", unique(treatment_2), "(wilcox)"))

obj2_rare %>%
  metacoder::filter_taxa(taxon_ranks == "o", supertaxa = TRUE) %>%
  metacoder::heat_tree(node_label = taxon_names,
          node_size = n_obs, # number of OTUs
          node_color = log2_median_ratio, # difference between groups
          node_color_interval = c(-max(abs(log2_median_ratio)), max(abs(log2_median_ratio))), # symmetric interval
          node_color_range = pal, # diverging colors
          node_size_axis_label = "OTU count",
          node_color_axis_label = "Log 2 ratio of median counts",
          layout = "davidson-harel", initial_layout = "reingold-tilford",
          title = paste(unique(treatment_1) ,"vs", unique(treatment_2), "(wilcox/rare)"))
```


```{r}
sapply(obj2$taxa, function(x) as.character(x$get_name())) %>% 
  as.data.frame %>%
  rename(taxa = ".")%>%
  rownames_to_column("taxon_id") -> taxo1_df2

sapply(obj2_rare$taxa, function(x)as.character(x$get_name())) %>% 
  as.data.frame %>%
  rename(taxa = ".")%>%
  rownames_to_column("taxon_id")-> taxo2_df2
table(taxo1_df2 == taxo2_df2)

obj2$data$diff_table %>%
  select(taxon_id, log2_median_ratio) %>%
  left_join(obj2_rare$data$diff_table %>%
            select(taxon_id, log2_median_ratio), by ="taxon_id" )%>% 
  ggplot()+
  geom_point(aes(log2_median_ratio.x, log2_median_ratio.y),size = 3, alpha =0.6)+
  geom_abline(intercept = 0, slope = 1, size = 2, col = "red")+
  labs(x =" log2(m1/m2) No rarefaction", y =" log2(m1/m2) rarefaction")+
  main_theme
  

```

#### Differential expression of replicated count data

```{r}
obj2 %>%
  metacoder::filter_taxa(taxon_ranks == "o", supertaxa = TRUE) %>%
  metacoder::heat_tree(node_label = taxon_names,
          node_size = n_obs, # number of OTUs
          node_color = logFC, # difference between groups
          node_color_interval = c(-max(abs(logFC)), max(abs(logFC))), # symmetric interval
          node_color_range = pal, # diverging colors
          node_size_axis_label = "OTU count",
          node_color_axis_label = "logFC",
          layout = "davidson-harel", initial_layout = "reingold-tilford",
          title = paste(unique(treatment_1) ,"vs", unique(treatment_2),"DGE"))

obj2_rare %>%
  metacoder::filter_taxa(taxon_ranks == "o", supertaxa = TRUE) %>%
  metacoder::heat_tree(node_label = taxon_names,
          node_size = n_obs, # number of OTUs
          node_color = logFC, # difference between groups
          node_color_interval = c(-max(abs(logFC)), max(abs(logFC))), # symmetric interval
          node_color_range = pal, # diverging colors
          node_size_axis_label = "OTU count",
          node_color_axis_label = "logFC",
          layout = "davidson-harel", initial_layout = "reingold-tilford",
          title = paste(unique(treatment_1) ,"vs", unique(treatment_2),"DGE rare"))


obj2$data$diff_analysis_table %>%
  select(taxon_id, logFC) %>%
  left_join(obj2_rare$data$diff_analysis_table %>%
            select(taxon_id, logFC), by ="taxon_id" ) %>% 
  ggplot()+
  geom_point(aes(logFC.x, logFC.y),size = 3, alpha =0.6)+
  geom_abline(intercept = 0, slope = 1, size = 2, col = "red")+
  labs(x =" log2(m1/m2) No rarefaction", y ="log2 (m1/m2) rarefaction")+
  main_theme
  
```

### Metabarcoding R1 vs Metagenomic R1

```{r}
meta_g_b_decond_table %>%
  make.tax.tree.comp(pair = c("R1_B","R1_G")) -> obj3
meta_g_b_decond_table_rare %>%
  make.tax.tree.comp(pair = c("R1_B","R1_G")) -> obj3_rare
```

####  Wilcox 2 by 2 comparison

```{r}

obj3 %>%
  metacoder::filter_taxa(taxon_ranks == "o", supertaxa = TRUE) %>%
  metacoder::heat_tree(node_label = taxon_names,
          node_size = n_obs, # number of OTUs
          node_color = log2_median_ratio, # difference between groups
          node_color_interval =  c(-max(abs(log2_median_ratio)), max(abs(log2_median_ratio))), # symmetric interval
          node_color_range = pal, # diverging colors
          node_size_axis_label = "OTU count",
          node_color_axis_label = "Log 2 ratio of median counts",
          layout = "davidson-harel", initial_layout = "reingold-tilford",
          title = paste(unique(treatment_1) ,"vs", unique(treatment_2)))

obj3_rare %>%
  metacoder::filter_taxa(taxon_ranks == "o", supertaxa = TRUE) %>%
  metacoder::heat_tree(node_label = taxon_names,
          node_size = n_obs, # number of OTUs
          node_color = log2_median_ratio, # difference between groups
          node_color_interval =  c(-max(abs(log2_median_ratio)), max(abs(log2_median_ratio))), # symmetric interval
          node_color_range = pal, # diverging colors
          node_size_axis_label = "OTU count",
          node_color_axis_label = "Log 2 ratio of median counts",
          layout = "davidson-harel", initial_layout = "reingold-tilford",
          title = paste(unique(treatment_1) ,"vs", unique(treatment_2)))
```

#### Differential expression of replicated count data

```{r}
obj3 %>%
  metacoder::filter_taxa(taxon_ranks == "o", supertaxa = TRUE) %>%
  metacoder::heat_tree(node_label = taxon_names,
          node_size = n_obs, # number of OTUs
          node_color = logFC, # difference between groups
          node_color_interval = c(-max(abs(logFC)), max(abs(logFC))), # symmetric interval
          node_color_range = pal, # diverging colors
          node_size_axis_label = "OTU count",
          node_color_axis_label = "logFC",
          layout = "davidson-harel", initial_layout = "reingold-tilford",
          title = paste(unique(treatment_1) ,"vs", unique(treatment_2),"DGE"))

obj3_rare %>%
  metacoder::filter_taxa(taxon_ranks == "o", supertaxa = TRUE) %>%
  metacoder::heat_tree(node_label = taxon_names,
          node_size = n_obs, # number of OTUs
          node_color = logFC, # difference between groups
          node_color_interval = c(-max(abs(logFC)), max(abs(logFC))), # symmetric interval
          node_color_range = pal, # diverging colors
          node_size_axis_label = "OTU count",
          node_color_axis_label = "logFC",
          layout = "davidson-harel", initial_layout = "reingold-tilford",
          title = paste(unique(treatment_1) ,"vs", unique(treatment_2),"DGE rare"))


obj3$data$diff_analysis_table %>%
  select(taxon_id, logFC) %>%
  left_join(obj3_rare$data$diff_analysis_table %>%
              select(taxon_id, logFC),
            by ="taxon_id" ) %>% 
  ggplot()+
  geom_point(aes(logFC.x, logFC.y),size = 3, alpha =0.6)+
  geom_abline(intercept = 0, slope = 1, size = 2, col = "red")+
  labs(x =" log2(m1/m2) No rarefaction", y ="log2 (m1/m2) rarefaction")+
  main_theme
```

### Metabarcoding R2 vs Metagenomic R2

```{r}
meta_g_b_decond_table %>%
  make.tax.tree.comp(pair = c("R2_B","R2_G")) -> obj4

meta_g_b_decond_table_rare %>%
  make.tax.tree.comp(pair = c("R2_B","R2_G")) -> obj4_rare

```

#### Wilcox 2 by 2 comparison

```{r}

obj4 %>%
  metacoder::filter_taxa(taxon_ranks == "o", supertaxa = TRUE) %>%
  metacoder::heat_tree(node_label = taxon_names,
          node_size = n_obs, # number of OTUs
          node_color = log2_median_ratio, # difference between groups
          node_color_interval =  c(-max(abs(log2_median_ratio)), max(abs(log2_median_ratio))), # symmetric interval
          node_color_range = pal, # diverging colors
          node_size_axis_label = "OTU count",
          node_color_axis_label = "Log 2 ratio of median counts",
          layout = "davidson-harel", initial_layout = "reingold-tilford",
          title = paste(unique(treatment_1) ,"vs", unique(treatment_2)))

obj4_rare %>%
  metacoder::filter_taxa(taxon_ranks == "o", supertaxa = TRUE) %>%
  metacoder::heat_tree(node_label = taxon_names,
          node_size = n_obs, # number of OTUs
          node_color = log2_median_ratio, # difference between groups
          node_color_interval =  c(-max(abs(log2_median_ratio)), max(abs(log2_median_ratio))), # symmetric interval
          node_color_range = pal, # diverging colors
          node_size_axis_label = "OTU count",
          node_color_axis_label = "Log 2 ratio of median counts",
          layout = "davidson-harel", initial_layout = "reingold-tilford",
          title = paste(unique(treatment_1) ,"vs", unique(treatment_2)))
```

#### Differential expression of replicated count data

```{r}
obj4 %>%
  metacoder::filter_taxa(taxon_ranks == "o", supertaxa = TRUE) %>%
  metacoder::heat_tree(node_label = taxon_names,
          node_size = n_obs, # number of OTUs
          node_color = logFC, # difference between groups
          node_color_interval = c(-max(abs(logFC)), max(abs(logFC))), # symmetric interval
          node_color_range = pal, # diverging colors
          node_size_axis_label = "OTU count",
          node_color_axis_label = "logFC",
          layout = "davidson-harel", initial_layout = "reingold-tilford",
          title = paste(unique(treatment_1) ,"vs", unique(treatment_2),"DGE"))

obj4_rare %>%
  metacoder::filter_taxa(taxon_ranks == "o", supertaxa = TRUE) %>%
  metacoder::heat_tree(node_label = taxon_names,
          node_size = n_obs, # number of OTUs
          node_color = logFC, # difference between groups
          node_color_interval = c(-max(abs(logFC)), max(abs(logFC))), # symmetric interval
          node_color_range = pal, # diverging colors
          node_size_axis_label = "OTU count",
          node_color_axis_label = "logFC",
          layout = "davidson-harel", initial_layout = "reingold-tilford",
          title = paste(unique(treatment_1) ,"vs", unique(treatment_2),"DGE rare"))


obj4$data$diff_analysis_table %>%
  select(taxon_id, logFC) %>%
  left_join(obj4_rare$data$diff_analysis_table %>%
            select(taxon_id, logFC), by ="taxon_id" ) %>% 
  ggplot()+
  geom_point(aes(logFC.x, logFC.y),size = 3, alpha =0.6)+
  geom_abline(intercept = 0, slope = 1, size = 2, col = "red")+
  labs(x =" log2 (m1/m2) No rarefaction", y ="log2 (m1/m2) rarefaction")+
  main_theme
```

Zero inflation might bias the variance of taxa abundance in each method thus scaling factors may be badly estimated

```{r}
data.frame(taxon_rank = sapply(names(obj1$taxa), function(x) obj1$taxa[[x]]$rank$name)) %>%
  rownames_to_column("taxon_id") -> rank_df

rank_df %>%
  left_join(obj1$data$tax_abund, by = "taxon_id" ) %>%
  tibble -> tax_abund
  
tax_abund %>%
    filter(taxon_rank == "k") %>%
    select(starts_with("BU"), taxon_id) %>%
    column_to_rownames("taxon_id") %>%
    as.matrix %>%
    hist(breaks = 50, main = "taxon_rank : kindom")

tax_abund %>%
    filter(taxon_rank == "p") %>%
    select(starts_with("BU"), taxon_id) %>%
    column_to_rownames("taxon_id") %>%
    as.matrix %>%
    hist(breaks = 50, main = "taxon_rank :phylum")

tax_abund %>%
    filter(taxon_rank == "c") %>%
    select(starts_with("BU"), taxon_id) %>%
    column_to_rownames("taxon_id") %>%
    as.matrix %>%
    hist(breaks = 50, main = "taxon_rank : class")

```

### Bayesian lognormal DGE zero inflated

Function that compute differential analysis (Bayesian) with zero inflated parameters and return an object compatible with metacoder
to plot it on the taxonomic comparison tree

```{r}
differential.analysis.Zif <- function(obj = obj,
                                        groups = c(79,79), #Indicates the size of each group samples
                                        tax_rank = c("k", "p", "c", "o", "f", "g"), # taxonomic rank in obj, comparison will be done for each rank
                                        nrAdapt=1000,
                                        nrBurnIn=1000,
                                        nrSamples=1000,
                                        nrThin = 3,
                                        chains = 2){
  ## Setup data out of metacoder object
  # Extract taxon rank associated to each Metacoder taxon_id
  data.frame(taxon_rank = sapply(names(obj$taxa), function(x) obj$taxa[[x]]$rank$name)) %>%
  rownames_to_column("taxon_id") -> rank_df
  
  # Join taxon rank and taxon total abundances data frame
  rank_df %>%
    left_join(obj$data$tax_abund, by = "taxon_id" ) %>%
  tibble -> tax_abund
  
  # Create sample data (data frame that indicate sample's condition/method belonging)
  data.frame(sample_ID = colnames(tax_abund %>%
                                    select(starts_with("BU"))
                                  )
             ) %>%
      mutate(method = str_extract(sample_ID, "R\\d_[A-Z+]$")) %>%
      column_to_rownames("sample_ID") -> sampledata
  ##
  
  c(rep(0,groups[1]),rep(1,groups[2])) -> group
  
  # indicator vector indicating which samples
  # belong to which group. Note this implementation requires
  # the two groups to be in consecutive order
  
  #MCMC run settings

  # Initialize output dataframe
  pair_wise_taxa_df <- data.frame()
  
  model_zif_list <- list()
  post_sample_list <-  list()

  for(t_rank in tax_rank){
    
    #add pseudo counts to any gene with zeros in one group
    tax_abund %>%
    filter(taxon_rank == t_rank) %>%
    select(starts_with("BU"), taxon_id) %>%
    column_to_rownames("taxon_id") %>%
    as.matrix -> Y
    str(Y)
    pseudoCounts <- 0
    for (i in 1:dim(Y)[1]){
      
      if(sum(Y[i,1:groups[1]])==0) { # add 1 to taxa with a 0 count across sample in group 1 
        Y[i,sample(1:groups[1],size=1)] <- 1
        pseudoCounts = pseudoCounts + 1
      }
      if(sum(Y[i,(1+groups[1]):(groups[2]+groups[1])])==0) {  # add 1 to taxa with a 0 count across sample in group 2
        Y[i,sample((1+groups[1]):(groups[2]+groups[1]),size=1)] <- 1
        
        pseudoCounts = pseudoCounts +1
      }
    }
      
  dim(Y) -> D 
  
  list(
    Y = Y,
    D = dim(Y),
    N = colSums(Y)/median(colSums(Y)),
    group = groups
  ) -> data_list
  
  list(
      alpha = rnorm(D[1]),
      beta = rnorm(D[1]),
      p = runif(n = D[1], min = 0, max = 1),
      binVar =  runif(D[1], 0.05, 0.1),
      phi = matrix(rnorm(D[1]*D[2]), ncol = D[2], nrow = D[1]),
      pi = Y*0,
      kappa = 1,
      eta = 1
    ) -> init_list

  rjags::jags.model("code/Zicompar.jags", data = data_list, inits = init_list, n.chains = chains, n.adapt = nrAdapt) -> model_zif
  
  update(model_zif,n.iter=nrBurnIn)

  rjags::coda.samples(model_zif,c("alpha", "beta", "p"),n.iter = nrSamples,thin=nrThin) -> post_sample

  coda::gelman.diag(post_sample, multivariate = F) -> param_convergence

  summary(post_sample) -> quantile_summary
  
  # post_sample %>% ggmcmc::ggs() -> postSamples_df
  # 
  # postSamples_df %>%
  # mutate(group = str_extract(Parameter,"(.+)\\[(\\d+)\\]", group = 1),
  #        taxa_id = str_extract(Parameter,"(.+)\\[(\\d+)\\]", group = 2)) -> postSamples_df

 param_convergence[[1]] %>%
   as.data.frame() %>%
   rownames_to_column("Parameter") %>%
   rename(point_est = "Point est.", upper_ci = "Upper C.I.") -> param_convergence_df

  quantile_summary$quantiles %>%
  data.frame() %>%
  rownames_to_column("Parameter") %>%
  filter(str_detect(Parameter,"beta")) %>%
  left_join(param_convergence_df, by = "Parameter") %>%
  mutate(signif = case_when(
    sign(X2.5.) != sign(X97.5.) ~ "NS",
    sign(X2.5.) == sign(X97.5.) ~ "sign"
    ),
    convergence = case_when(
    upper_ci < 1.2 ~ T,
    upper_ci > 1.2 ~ F
    )
  ) %>%
  select(beta_lower = X2.5., beta_upper = X97.5., beta = X50., signif, convergence, point_est, upper_ci) -> resultdf


  rownames(resultdf) = rownames(Y)
  resultdf %>%
    rownames_to_column("taxon_id") -> resultdf
  
  pair_wise_taxa_df %>%
    bind_rows(resultdf)%>%
    tibble() -> pair_wise_taxa_df
  
  model_zif_list = c(model_zif_list, list(model_zif))
  post_sample_list = c(post_sample_list, list(post_sample))

  }

  return(list(model_zif, post_sample, pair_wise_taxa_df))
}
```

Function for rarefied data. Same function than the previous one but the Bayesian model doesn't contain a standardization by taxa abundance sum per sample (used to control sampling effort)

```{r}
differential.analysis.Zif.raref <- function(obj = obj,
                                        groups = c(79,79), #Indicates the size of each group samples
                                        tax_rank = c("k", "p", "c", "o", "f", "g"), # taxonomic rank in obj, comparison will be done for each rank
                                        nrAdapt=1000,
                                        nrBurnIn=1000,
                                        nrSamples=1000,
                                        nrThin = 3,
                                        chains = 2){
  ## Setup data out of metacoder object
  # Extract taxon rank associated to each Metacoder taxon_id
  data.frame(taxon_rank = sapply(names(obj$taxa), function(x) obj$taxa[[x]]$rank$name)) %>%
  rownames_to_column("taxon_id") -> rank_df
  
  # Join taxon rank and taxon total abundances data frame
  rank_df %>%
    left_join(obj$data$tax_abund, by = "taxon_id" ) %>%
  tibble -> tax_abund
  
  # Create sample data (data frame that indicate sample's condition/method belonging)
  data.frame(sample_ID = colnames(tax_abund %>%
                                    select(starts_with("BU"))
                                  )
             ) %>%
      mutate(method = str_extract(sample_ID, "R\\d_[A-Z+]$")) %>%
      column_to_rownames("sample_ID") -> sampledata
  ##
  
  c(rep(0,groups[1]),rep(1,groups[2])) -> group
  
  # indicator vector indicating which samples
  # belong to which group. Note this implementation requires
  # the two groups to be in consecutive order
  
  #MCMC run settings

  # Initialize output dataframe
  pair_wise_taxa_df <- data.frame()
  
  model_zif_list <- list()
  post_sample_list <-  list()
  for(t_rank in tax_rank){
    
    tax_abund %>%
    filter(taxon_rank == t_rank) %>%
    select(starts_with("BU"), taxon_id) %>%
    column_to_rownames("taxon_id") %>%
    as.matrix -> Y
  
  dim(Y) -> D     
     
  list(
    Y = Y,
    D = dim(Y),
    group = groups
  ) -> data_list
  
  list(
      alpha = rnorm(D[1]),
      beta = rnorm(D[1]),
      p = runif(n = D[1], min = 0, max = 1),
      binVar =  runif(D[1], 0.05, 0.1),
      phi = matrix(rnorm(D[1]*D[2]), ncol = D[2], nrow = D[1]),
      pi = Y*0,
      kappa = 1,
      eta = 1
    ) -> init_list
    
    rjags::jags.model("code/Zicompar_rarefaction.jags", data = data_list, inits = init_list, n.chains = chains, n.adapt = nrAdapt) -> model_zif
    
    update(model_zif,n.iter=nrBurnIn)
    
    rjags::coda.samples(model_zif,c("alpha", "beta", "p"),n.iter = nrSamples,thin=nrThin) -> post_sample
    
    coda::gelman.diag(post_sample, multivariate = F) -> param_convergence
    
    summary(post_sample) -> quantile_summary
    
    # post_sample %>% ggmcmc::ggs() -> postSamples_df
    # 
    # postSamples_df %>%
    # mutate(group = str_extract(Parameter,"(.+)\\[(\\d+)\\]", group = 1),
    #        taxa_id = str_extract(Parameter,"(.+)\\[(\\d+)\\]", group = 2)) -> postSamples_df
    
    param_convergence[[1]] %>%
     as.data.frame() %>%
     rownames_to_column("Parameter") %>%
     rename(point_est = "Point est.", upper_ci = "Upper C.I.") -> param_convergence_df
    
    quantile_summary$quantiles %>%
    data.frame() %>%
    rownames_to_column("Parameter") %>%
    filter(str_detect(Parameter,"beta")) %>%
    left_join(param_convergence_df, by = "Parameter") %>%
    mutate(signif = case_when(
      sign(X2.5.) != sign(X97.5.) ~ "NS",
      sign(X2.5.) == sign(X97.5.) ~ "sign"
      ),
      convergence = case_when(
      upper_ci < 1.2 ~ T,
      upper_ci > 1.2 ~ F
      )
    ) %>%
    select(beta_lower = X2.5., beta_upper = X97.5., beta = X50., signif, convergence, point_est, upper_ci) -> resultdf
    
    
    rownames(resultdf) = rownames(Y)
    resultdf %>%
      rownames_to_column("taxon_id") -> resultdf
    
    pair_wise_taxa_df %>%
      bind_rows(resultdf)%>%
      tibble() -> pair_wise_taxa_df
    
    model_zif_list = c(model_zif_list, list(model_zif))
    post_sample_list = c(post_sample_list, list(post_sample))

  }

  return(list(model_zif, post_sample, pair_wise_taxa_df))
}
```

### Function to compute bayesian model in parallel

```{r}
obj1$data$tax_data
obj1_rare$data$tax_data
differential.analysis.Zif.raref(obj = obj1_rare, groups = c(79,79),
                                 tax_rank = c("k", "p", "c", "o"),
                                 nrAdapt=100,
                                 nrBurnIn=100,
                                 nrSamples=50,
                                 chains = 5,
                                 nrThin = 5)


parallel.DA.Zif <- function(obj , groups, tax_rank, data_type,
                            nrAdapt, nrBurnIn, nrSamples, chains, nrThin){
  if(data_type == "raref"){
    differential.analysis.Zif.raref(obj = obj, groups = c(79,79),
                                 tax_rank = tax_rank,
                                 nrAdapt=nrAdapt,
                                 nrBurnIn=nrBurnIn,
                                 nrSamples=nrSamples,
                                 chains = chains,
                                 nrThin = nrThin)
  }else{
    differential.analysis.Zif(obj = obj, groups = c(79,79),
                                 tax_rank = tax_rank,
                                 nrAdapt=nrAdapt,
                                 nrBurnIn=nrBurnIn,
                                 nrSamples=nrSamples,
                                 chains = chains,
                                 nrThin = nrThin)
  }
}

```

### Compute bayesian model

```{r}

if(file.exists("outputs/diff_analysis_list_zif.RData")){
  load("outputs/diff_analysis_list_zif.RData")
}else{
  print("go get a coffee.... it will probably take the day")
  cl <- parallel::makeCluster(8)
  doParallel::registerDoParallel(cl)
  diff_analysis_list_zif <- foreach::foreach(tree_obj = iter(list(obj1, obj1_rare ,obj2, obj2_rare,obj3, obj3_rare,obj4, obj4_rare)),
                                           args = c("no_raref","raref", "no_raref","raref", "no_raref","raref", "no_raref","raref"),
                                           .packages = c("rjags", "tidyverse", "coda", "ggmcmc"),
                                           .verbose = T) %dopar%
    parallel.DA.Zif(obj = tree_obj, groups = c(79,79),
                  data_type = args,
                    tax_rank =  c("k", "p", "c", "o"),
                    nrAdapt=1000,
                    nrBurnIn=1000,
                    nrSamples=5000,
                    chains = 5,
                    nrThin = 5)
    parallel::stopCluster(cl)
    
    save(diff_analysis_list_zif,file = "outputs/diff_analysis_list_zif.RData")
    save.image()
}
  


```


```{r}
for(i in 1:8){
  names(  diff_analysis_list_zif[[i]][[3]])[1] <- "taxon_id"
  
  diff_analysis_list_zif[[i]][[3]] %>%
    mutate(beta_sign = case_when(
      (signif == "NS")  ~ 0,
      !convergence  ~ 0,
     (signif == "sign") & convergence  ~ beta
      )
     ) -> diff_analysis_list_zif[[i]][[3]]
  }

obj1$data$diff_analysis_table_zif <- diff_analysis_list_zif[[1]][[3]] 
obj1_rare$data$diff_analysis_table_zif <- diff_analysis_list_zif[[2]][[3]]

obj2$data$diff_analysis_table_zif <- diff_analysis_list_zif[[3]][[3]] 
obj2_rare$data$diff_analysis_table_zif <- diff_analysis_list_zif[[4]][[3]]

obj3$data$diff_analysis_table_zif <- diff_analysis_list_zif[[5]][[3]] 
obj3_rare$data$diff_analysis_table_zif <- diff_analysis_list_zif[[6]][[3]]

obj4$data$diff_analysis_table_zif <- diff_analysis_list_zif[[7]][[3]] 
obj4_rare$data$diff_analysis_table_zif <- diff_analysis_list_zif[[8]][[3]]
```

### R1_B vs R2_B

```{r}

obj1%>%
  metacoder::filter_taxa(taxon_ranks == "o", supertaxa = TRUE)%>%
  metacoder::heat_tree(node_label = taxon_names,
          node_size = n_obs, # number of OTUs
          node_color = beta_sign, # difference between groups
          node_color_interval = c(-max(abs(beta_sign)), max(abs(beta_sign))), # symmetric interval
          node_color_range = pal, # diverging colors
          node_size_axis_label = "OTU count",
          node_color_axis_label = "logFC (beta)",
          layout = "davidson-harel", initial_layout = "reingold-tilford",
          title = paste(unique(treatment_1) ,"vs", unique(treatment_2),"bayes"))

# without accounting for significance and convergence
obj1$data$diff_analysis_table_zif %>%
  select(taxon_id, beta) %>%
  left_join(obj1_rare$data$diff_analysis_table_zif %>%
            select(taxon_id, beta), by ="taxon_id" )%>% 
  ggplot()+
  geom_point(aes(beta.x, beta.y),size = 3, alpha =0.6)+
  geom_abline(intercept = 0, slope = 1, linewidth = 2, col = "red")+
  labs(x =" beta(logFC) no rarefaction", y ="beta(logFC) rarefaction")+
  main_theme

# with significance and convergence
obj1$data$diff_analysis_table %>%
  select(taxon_id, logFC) %>%
  left_join(obj1$data$diff_analysis_table_zif %>%
            select(taxon_id, beta_sign), by ="taxon_id" ) %>% 
  ggplot()+
  geom_point(aes(logFC, -beta_sign),size = 3, alpha =0.6)+
  geom_abline(intercept = 0, slope = 1, linewidth = 2, col = "red")+
  labs(x =" logFC No rarefaction", y ="Zif logFC no rarefaction")+
  main_theme
save.image()
```

### R1_G vs R2_G

```{r}
obj2%>%
  metacoder::filter_taxa(taxon_ranks == "o", supertaxa = TRUE)%>%
  metacoder::heat_tree(node_label = taxon_names,
          node_size = n_obs, # number of OTUs
          node_color = beta_sign, # difference between groups
          node_color_interval = c(-max(abs(beta_sign)), max(abs(beta_sign))), # symmetric interval
          node_color_range = pal, # diverging colors
          node_size_axis_label = "OTU count",
          node_color_axis_label = "logFC (beta)",
          layout = "davidson-harel", initial_layout = "reingold-tilford",
          title = paste(unique(treatment_1) ,"vs", unique(treatment_2),"bayes"))
# without accounting for significance and convergence
obj2$data$diff_analysis_table_zif %>%
  select(taxon_id, beta) %>%
  left_join(obj2_rare$data$diff_analysis_table_zif %>%
            select(taxon_id, beta), by ="taxon_id" )%>% 
  ggplot()+
  geom_point(aes(beta.x, beta.y),size = 3, alpha =0.6)+
  geom_abline(intercept = 0, slope = 1, linewidth = 2, col = "red")+
  labs(x =" beta(logFC) no rarefaction", y ="beta(logFC) rarefaction")+
  main_theme

# with significance and convergence
obj2$data$diff_analysis_table %>%
  select(taxon_id, logFC) %>%
  left_join(obj2$data$diff_analysis_table_zif %>%
            select(taxon_id, beta_sign), by ="taxon_id" ) %>% 
  ggplot()+
  geom_point(aes(logFC, -beta_sign),size = 3, alpha =0.6)+
  geom_abline(intercept = 0, slope = 1, linewidth = 2, col = "red")+
  labs(x =" logFC No rarefaction", y ="Zif logFC no rarefaction")+
  main_theme
```

### R1_B vs R1_G

```{r}
obj3%>%
  metacoder::filter_taxa(taxon_ranks == "o", supertaxa = TRUE)%>%
  metacoder::heat_tree(node_label = taxon_names,
          node_size = n_obs, # number of OTUs
          node_color = beta_sign, # difference between groups
          node_color_interval = c(-max(abs(beta_sign)), max(abs(beta_sign))), # symmetric interval
          node_color_range = pal, # diverging colors
          node_size_axis_label = "OTU count",
          node_color_axis_label = "logFC (beta)",
          layout = "davidson-harel", initial_layout = "reingold-tilford",
          title = paste(unique(treatment_1) ,"vs", unique(treatment_2),"bayes"))
# without accounting for significance and convergence
obj3$data$diff_analysis_table_zif %>%
  select(taxon_id, beta) %>%
  left_join(obj3_rare$data$diff_analysis_table_zif %>%
            select(taxon_id, beta), by ="taxon_id" )%>% 
  ggplot()+
  geom_point(aes(beta.x, beta.y),size = 3, alpha =0.6)+
  geom_abline(intercept = 0, slope = 1, linewidth = 2, col = "red")+
  labs(x =" beta(logFC) no rarefaction", y ="beta(logFC) rarefaction")+
  main_theme

# with significance and convergence
obj3$data$diff_analysis_table %>%
  select(taxon_id, logFC) %>%
  left_join(obj3$data$diff_analysis_table_zif %>%
            select(taxon_id, beta_sign), by ="taxon_id" ) %>% 
  ggplot()+
  geom_point(aes(logFC, -beta_sign),size = 3, alpha =0.6)+
  geom_abline(intercept = 0, slope = 1, linewidth = 2, col = "red")+
  labs(x =" logFC No rarefaction", y ="Zif logFC no rarefaction")+
  main_theme
```

### R2_B vs R2_G

```{r}
obj4%>%
  metacoder::filter_taxa(taxon_ranks == "o", supertaxa = TRUE)%>%
  metacoder::heat_tree(node_label = taxon_names,
          node_size = n_obs, # number of OTUs
          node_color = beta_sign, # difference between groups
          node_color_interval = c(-max(abs(beta_sign)), max(abs(beta_sign))), # symmetric interval
          node_color_range = pal, # diverging colors
          node_size_axis_label = "OTU count",
          node_color_axis_label = "logFC (beta)",
          layout = "davidson-harel", initial_layout = "reingold-tilford",
          title = paste(unique(treatment_1) ,"vs", unique(treatment_2),"bayes"))
# without accounting for significance and convergence
obj4$data$diff_analysis_table_zif %>%
  select(taxon_id, beta) %>%
  left_join(obj4_rare$data$diff_analysis_table_zif %>%
            select(taxon_id, beta), by ="taxon_id" )%>% 
  ggplot()+
  geom_point(aes(beta.x, beta.y),size = 3, alpha =0.6)+
  geom_abline(intercept = 0, slope = 1, linewidth = 2, col = "red")+
  labs(x =" beta(logFC) no rarefaction", y ="beta(logFC) rarefaction")+
  main_theme

# with significance and convergence
obj4$data$diff_analysis_table %>%
  select(taxon_id, logFC) %>%
  left_join(obj4$data$diff_analysis_table_zif %>%
            select(taxon_id, beta_sign), by ="taxon_id" ) %>% 
  ggplot()+
  geom_point(aes(logFC, -beta_sign),size = 3, alpha =0.6)+
  geom_abline(intercept = 0, slope = 1, linewidth = 2, col = "red")+
  labs(x =" logFC No rarefaction", y ="Zif logFC no rarefaction")+
  main_theme
```


## Annexe

This Bayesian model use poisson log normal distribution
Key Characteristics

Poisson Component: The Poisson component represents the underlying count process, where the mean and variance are linked through the parameter  (lambda).
Lognormal Component: The log-normal component models the uncertainty in the Poisson parameter , introducing variability and skewness to the distribution.
Compound Distribution: The Poisson-log-normal distribution is a compound distribution, where the Poisson distribution is conditioned on the log-normal distribution of .

```{r}
# log normal
exp(rnorm(1000)) -> log_norm
hist(log_norm, breaks =100)

# normal
hist(log(log_norm), breaks =100)

# poisson log normal
rpois(n = 1000,lambda = exp(rnorm(1000)))%>%
  hist(breaks =100)
```

